---
title: "Breast Cancer Survival Analysis: EDA"
format: 
  html:
    embed-resources: true
    date-modified: today
    code-fold: true
    toc: true

---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(rafalib)
library(kableExtra)
library(gridExtra)
library(knitr)

library(survival)
library(survminer)
library(car)  # for vif()


data <- read.csv("../data/data_clean.csv")

# overall_survival is not mutated, as it needs to be numeric for survival analysis

# merge 1-2, 3-5, above 5 -- for consistency

# pairwise correlation for interaction effect ? 

# check the line at 1


data <- data |>
    mutate(lymph_nodes_group = case_when(
    lymph_nodes_examined_positive == 0 ~ "0",
    lymph_nodes_examined_positive == 1 ~ "1",
    lymph_nodes_examined_positive == 2 ~ "2",
    lymph_nodes_examined_positive <= 5 ~ "3-5",
    lymph_nodes_examined_positive > 5 ~ ">5"
  )) |>
  mutate(
    tumor_stage = as.factor(tumor_stage),
    her2_status = as.factor(her2_status),
    er_status = as.factor(er_status),
    pr_status = as.factor(pr_status),
    her2_status_measured_by_snp6 = as.factor(her2_status_measured_by_snp6),
    death_from_cancer = as.factor(death_from_cancer),
    neoplasm_histologic_grade = as.factor(neoplasm_histologic_grade),
    lymph_nodes_group = as.factor(lymph_nodes_group)
  ) |> filter(tumor_stage != "0", tumor_stage != "4")
  

data <- droplevels(data)

table(data$lymph_nodes_examined_positive)
table(data$lymph_nodes_group)
```



# 1. KM Survival Curves

## 1.1. Assumptions

* Events are independent of each other. 

* Right-censoring is uninformative.

* Event times are exact.

## 1.2. Plotting survival curves

### 1.2.1. Overall curve

Fit a Kaplan-Meier Survival curve for all the data:

```{r warning=FALSE}

km <- survfit(Surv(data$overall_survival_months, data$overall_survival) ~ 1)
km

survival_table <- summary(km)

ggsurvplot(km,
           data = data,              
           conf.int = FALSE,
           xlab = "Time (months)",
           ylab = "Survival Probability",
           surv.median.line = "hv",
           xlim = c(0, 350),        # show only first 60 months
           break.time.by = 50,      # tick marks every 12 months
           legend = "none")

```
* There were 1,400 patients in total.

* 610 experienced the event (death). The rest (790) were censored.

* The median survival time is 200 months. After that, the survival probability drops below 0.5.

* The 95% confidence interval for the median survival time is 194 to 211 months.

### 1.2.2. Curves by Tumour Stage


```{r}
fit_stage <- survfit(Surv(data$overall_survival_months, data$overall_survival) ~ tumor_stage, data = data)

# KM plot with log-rank p-value
ggsurvplot(fit_stage, data = data, pval.method = TRUE, pval = TRUE, 
           conf.int = TRUE,
           legend.title = "Tumor Stage",
           legend.labs = levels(data$tumor_stage))
```

Stratified by HER2 status:

```{r}
# HER2 status
fit_tumor_by_her2 <- survfit(Surv(overall_survival_months, overall_survival) ~ tumor_stage + her2_status, data = data)

ggsurvplot_facet(
  fit_tumor_by_her2,
  data = data,
  facet.by = "her2_status",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "Tumour Stage",
  legend.labs = levels(data$tumor_stage),
  palette = "Dark2"
)
```

Stratified by ER status:

```{r}
# ER status
fit_tumor_by_er <- survfit(Surv(overall_survival_months, overall_survival) ~ tumor_stage + er_status, data = data)

ggsurvplot_facet(
  fit_tumor_by_er,
  data = data,
  facet.by = "er_status",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "Tumour Stage",
  legend.labs = levels(data$tumor_stage),
  palette = "Dark2"
)
```

Stratified by lymph node size:

```{r}
# lymph node size (levels)

fit_tumor_by_lymph <- survfit(Surv(overall_survival_months, overall_survival) ~ tumor_stage + lymph_nodes_group, data = data)

ggsurvplot_facet(
  fit_tumor_by_lymph,
  data = data,
  facet.by = "lymph_nodes_group",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "Tumour Stage",
  legend.labs = levels(data$tumor_stage),
  palette = "Dark2"
)

```

### 1.2.3. Curves by HER2 Status

```{r}
fit_her2 <- survfit(Surv(data$overall_survival_months, data$overall_survival) ~ her2_status, data = data)

# KM plot with log-rank p-value
ggsurvplot(fit_her2, data = data, pval.method = TRUE, pval = TRUE, 
           conf.int = TRUE,
           legend.title = "HER2 Status",
           legend.labs = levels(data$her2_status))
```

Stratified by tumour stage:

```{r}
fit_her2_by_tumor <- survfit(Surv(overall_survival_months, overall_survival) ~ her2_status + tumor_stage, data = data)

ggsurvplot_facet(
  fit_her2_by_tumor,
  data = data,
  facet.by = "tumor_stage",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "HER2 Status",
  legend.labs = levels(data$her2_status),
  palette = "Dark2"
)
```


Stratified by ER status:


```{r}
fit_her2_by_er <- survfit(Surv(overall_survival_months, overall_survival) ~ her2_status + er_status, data = data)

ggsurvplot_facet(
  fit_her2_by_er,
  data = data,
  facet.by = "er_status",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "HER2 Status",
  legend.labs = levels(data$her2_status),
  palette = "Dark2"
)


```


Stratified by lymph node size:


```{r}
fit_her2_by_lymph <- survfit(Surv(overall_survival_months, overall_survival) ~ her2_status + lymph_nodes_group, data = data)

ggsurvplot_facet(
  fit_her2_by_lymph,
  data = data,
  facet.by = "lymph_nodes_group",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "HER2 Status",
  legend.labs = levels(data$her2_status),
  palette = "Dark2"
)


```


### 1.2.4. Curves by ER Status


```{r}
fit_er <- survfit(Surv(data$overall_survival_months, data$overall_survival) ~ er_status, data = data)

# KM plot with log-rank p-value
ggsurvplot(fit_er, data = data, pval.method = TRUE, pval = TRUE, 
           conf.int = TRUE,
           legend.title = "ER Status",
           legend.labs = levels(data$er_status))
```

Stratified by tumour stage:


```{r}
fit_er_by_tumor <- survfit(Surv(overall_survival_months, overall_survival) ~ er_status + tumor_stage, data = data)


ggsurvplot_facet(
  fit_er_by_tumor,
  data = data,
  facet.by = "tumor_stage",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "ER Status",
  legend.labs = levels(data$er_status),
)
```

Stratified by HER2 status:


```{r}
fit_er_by_her2 <- survfit(Surv(overall_survival_months, overall_survival) ~ er_status + her2_status, data = data)

ggsurvplot_facet(
  fit_er_by_her2,
  data = data,
  facet.by = "her2_status",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "ER Status",
  legend.labs = levels(data$er_status),
  palette = "Dark2"
)


```


Stratified by lymph node size:


```{r}
fit_er_by_lymph <- survfit(Surv(overall_survival_months, overall_survival) ~ er_status + lymph_nodes_group, data = data)

ggsurvplot_facet(
  fit_er_by_lymph,
  data = data,
  facet.by = "lymph_nodes_group",
  pval = TRUE,
  pval.method = TRUE,
  conf.int = TRUE,
  legend.title = "ER Status",
  legend.labs = levels(data$er_status),
  palette = "Dark2"
)


```

# 2. Cox Proportional Hazards (PH) Model: Assumptions

Along with the assumptions for KM above in 1.1, we check for the following additional assumptions:

## 2.1. Constant Hazard Ratios (HR) are constant

The hazard ratio needs to be constant over time (e.g if the hazard at a specific timepoint for someone with tumour stage 2 is **double** that of one with stage 1, this ratio needs to stay constant over all time i.e., not dependent on time). If it isn't, we need to introduce the covariate as a time-dependent covariate in our model.


### 2.1.1 Log Minus Log (LML) check

We can use LML graphs to check if this assumption is met. 

Hazard ratios between two groups are constant over time. If true, then log minus long (log(-log(S))) graphs should be parallel (and not cross over):


Tumour stage: 

```{r}

# tumor stage

fit_stage_log <- survfit(Surv(overall_survival_months, overall_survival) ~ tumor_stage, data = data)


plot(fit_stage_log,
     fun = "cloglog",  # log(-log(S))
     xscale = 2,       # log10 scale on x-axis
     col = 1:2,
     lty = 1:2)

```

HER2 status:

```{r}
# er2 status

fit_her2_log <- survfit(Surv(overall_survival_months, overall_survival) ~ her2_status, data = data)


plot(fit_her2_log,
     fun = "cloglog",  # log(-log(S))
     xscale = 2,       # log10 scale on x-axis
     col = 1:2,
     lty = 1:2)

```

ER status: 

```{r}
# ER status

fit_er_log <- survfit(Surv(overall_survival_months, overall_survival) ~ er_status, data = data)


plot(fit_er_log,
     fun = "cloglog",  # log(-log(S))
     xscale = 2,       # log10 scale on x-axis
     col = 1:2,
     lty = 1:2)

```



### 2.1.2 "Fit the model and check for residuals" check

We can go ahead and fit the Cox model. Running cox.zph(fit_ph) can verify if the model assumption of proportional hazards is met. We can also look at the below residuals-- if there is no trend (horizontal), then the assumptions are likely met.

LML plots are not parallel which imply constant HR assumptions were violated but I fitted the Cox PH model anyway and the residuals were horizontal (and p > 0.05 for all variables), so most likely it was because taking all covariates in account together results in PH ration being constant-- because the crossing of the LML plots from before can be due to a covariate missing as well as time-dependency.


```{r}

fit_her2_log <- survfit(Surv(overall_survival_months, overall_survival) ~ her2_status, data = data)

fit_ph <- coxph(Surv(overall_survival_months, overall_survival) ~ tumor_stage + age_at_diagnosis + her2_status + er_status + lymph_nodes_examined_positive + tumor_size, data = data)

test.ph <- cox.zph(fit_ph)
print(test.ph)
plot(test.ph)




```

## 2.2. Linearity

If a variable’s effect is not linear on the log hazard scale, treating it as if it were can introduce bias into the model, causing some variables to appear more influential than they truly are.

We use Martingale residuals to check the linear assumption. If there is a non-linear trend, then a transformation of the data is required:

```{r}

fit_ph <- coxph(Surv(overall_survival_months, overall_survival) ~ tumor_stage + age_at_diagnosis + her2_status + er_status + lymph_nodes_examined_positive + tumor_size, data = data)

martingale_resid <- residuals(fit_ph, type = "martingale")

# Age at diagnosis
plot(data$age_at_diagnosis, martingale_resid)

# Lymph nodes examined positive
plot(data$lymph_nodes_examined_positive, martingale_resid)

# Tumor size
plot(data$tumor_size, martingale_resid)


```
We need to log transform lymph nodes and tumor size:

```{r}

data$log_lymph_nodes <- log(data$lymph_nodes_examined_positive + 1)

fit_ph_log <- coxph(Surv(overall_survival_months, overall_survival) ~ tumor_stage + age_at_diagnosis + her2_status + er_status + log_lymph_nodes + tumor_size, data = data)

martingale_resid <- residuals(fit_ph_log, type = "martingale")
plot(data$log_lymph_nodes, martingale_resid,
     xlab = "log(lymph_nodes_examined_positive + 1)",
     ylab = "Martingale residuals",
     main = "Checking linearity after log transformation")


data$log_tumor <- log(data$tumor_size + 1)

fit_ph_log_tumor <- coxph(Surv(overall_survival_months, overall_survival) ~ tumor_stage + age_at_diagnosis + her2_status + er_status + log_lymph_nodes + log_tumor, data = data)

martingale_resid <- residuals(fit_ph_log_tumor, type = "martingale")
plot(data$log_tumor, martingale_resid,
     xlab = "log(tumor_size)",
     ylab = "Martingale residuals",
     main = "Checking linearity after log transformation")


fit_lin_node  <- coxph(Surv(overall_survival_months, overall_survival) ~ log(lymph_nodes_examined_positive+1) + tumor_stage + age_at_diagnosis + her2_status + er_status + log_tumor, data=data)

fit_spl_node  <- coxph(Surv(overall_survival_months, overall_survival) ~ pspline(log(lymph_nodes_examined_positive+1)) + tumor_stage + age_at_diagnosis + her2_status + er_status + log_tumor, data=data)

anova(fit_lin_node, fit_spl_node, test="Chisq")
 


fit_lin_stage  <- coxph(Surv(overall_survival_months, overall_survival) ~ log_lymph_nodes + tumor_stage + age_at_diagnosis + her2_status + er_status + log_tumor, data=data)

fit_spl_stage  <- coxph(Surv(overall_survival_months, overall_survival) ~ pspline(log(tumor_size+1)) + log_lymph_nodes + tumor_stage + age_at_diagnosis + her2_status + er_status, data=data)

anova(fit_lin_stage, fit_spl_stage, test="Chisq")

```


## 2.3. Multicollinearity 

If two variables are highly correlated, it could introduce further variance into the model.

All VIF (Variance Inflation Factor) values are below 5, indicating no serious multicollinearity.

```{r warning=FALSE}

vif(fit_ph)

```

## 2.4. Interaction

```{r}


cox_interaction <- coxph(Surv(overall_survival_months, overall_survival) ~ (tumor_stage + age_at_diagnosis + her2_status + er_status + lymph_nodes_examined_positive + tumor_size)^2, data = data)

round(
  cbind("AHR"      = exp(summary(cox_interaction)$coef[, "coef"]),
                     exp(confint(cox_interaction)),
        "p-value"  = summary(cox_interaction)$coef[, "Pr(>|z|)"])
  , 3)


```
The interaction between her2_status and tumor_size (p=.042), and between er_status and lymph_nodes_examined positive (p=.038) are significant; the other variables are fine. This means we need to include an interaction term between the two pairs, re-fit the model, and re-check previous assumptions.

```{r}
data <- data |>
  mutate(
    log_tumor_size = log(tumor_size),
    log_nodes_positive = log(lymph_nodes_examined_positive + 1) # add 1 to avoid log(0)
  )

fit_log_int <- coxph(Surv(overall_survival_months, overall_survival) ~ 
                   tumor_stage + age_at_diagnosis +
                   her2_status * log_tumor_size +
                   er_status * log_nodes_positive,
                 data = data)


test.ph_int <- cox.zph(fit_log_int)
plot(test.ph_int)
```

Hm... the plot for ER_status curves up in the end a bit after 240 months. Print test.ph_int to see if there are any formal statistically significant variations over time between variables:

```{r}
print(test.ph_int)

```

There are no statistically significant differences in overall variation in time, (p >0.05), even though the er_status curves upwards. Therefore, the PH hazards assumption is valid.


```{r}
martingale_resid <- residuals(fit_log_int, type = "martingale")

# Age at diagnosis
plot(data$age_at_diagnosis, martingale_resid)

# Lymph nodes examined positive
plot(data$log_nodes_positive, martingale_resid)

# Tumor size
plot(data$log_tumor_size, martingale_resid)

```

All the above residuals look good! Seem to validify linear assumption after log transformation of lymph nodes and tumor size :)

```{r}
vif(fit_log_int)

```

The slight multicollinearity between her2_status:log_tumor_size is acceptable because the interaction is both statistically significant and biologically meaningful (i.e. the slope for tumor size isn’t the same in HER2+ and HER2− patients.)

```{r}
summary(fit_log_int)
```

* As HR increases with increasing tumor stage (9% for Stage 2 and 30% for Stage 3 compared to S1), but the differences are not significant. 

* Tumour size and her2 status are not significant by itself.

```{r}
library(dplyr)
library(broom)

# Get tidy summary WITHOUT exponentiation
tidy_fit <- broom::tidy(fit_log_int, exponentiate = FALSE, conf.int = TRUE)

# Manually compute HR and its CI
summary_table <- tidy_fit |>
  mutate(
    HR = exp(estimate),
    conf.low.HR = exp(conf.low),
    conf.high.HR = exp(conf.high),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE ~ ""
    ),
    effect_interpretation = case_when(
      HR > 1 ~ "Higher hazard (↑ risk)",
      HR < 1 ~ "Lower hazard (↓ risk)",
      TRUE ~ "No effect"
    )
  ) |>
  select(term, estimate, HR, conf.low.HR, conf.high.HR, p.value,
         significance, effect_interpretation) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

kable(
  summary_table,
  caption = "Cox Proportional Hazards Model Summary",
  align = "lccccccc",
  format = "markdown"
)

```


